### 2017-06-25 Manuscript/Response_to_reviewers.Rmd

RESPONSE: We now clarify (lines 227-232) that the territory quality estimates that we use in our statistical models are taken from the field period in which the telomere sample was taken. The same applies to insect abundance, group size etc. We feel that an in-depth explanation of how and why territory quality varies over space and time is outside the scope of this manuscript, especially given i) the word limit, and ii) the fact that territory quality is not related to RTL. However we do now include a reference to a recent book chapter (Komdeur et al. 2016) which explains how and why territory quality varies on Cousin Island. Our updated section on territory quality and indect abundance is now on lines 142-150:

@@

\begin{adjustwidth}{2cm}{2cm}

\textit{"Where territory quality estimates were not available for a specific year we used the average value for that territory across years (Hammers et al. 2013; see Komdeur et al. 2016 for an explanation of how territory quality varies on Cousin Island). Cousin is subject to considerable intra- and inter-annual variation in rainfall and, consequently, insect availability (Komdeur et al. 2016). Such island-wide temporal variation may override the effects of variation in individual territory quality across the island. As an estimate of seasonal variation in food availability, we calculated an index of the abundance of insects across the entire island during each main breeding season (referred to hereafter as ‘insect abundance’). This index is calculated as the mean number of insects found per unit leaf area over all monthly surveys carried out on the island in a main breeding season."}

@@

---

\bigskip

**L 165: “time point t + 1” Is the time span between repeated samples always the same? If not, please add a sentence to make it clear. Are samples taken in the same month every year?**

RESPONSE: Samples are taken as and when we can catch the birds, which is not always in the same month, but samples are generally taken at the same time of year. We now clarify this on lines 177-178:

@@

\begin{adjustwidth}{2cm}{2cm}

\textit{"Individuals were not always caught in the same month, but were generally caught within a 3 month breeding season window."}

@@

---

\bigskip

**Some more information on the sampled individuals your study is based on would be helpful and could be added to the supplement. E.g. cohort sizes.**

RESPONSE: We have now added a table of cohort sizes to the supplement (Table S1), and refer to this in the main text (e.g. on Line 126).

@@

---

\bigskip

**L 234: Please provide intra-plate repeatabilities as well. Do you have repeats that are not directly next to each other on the plate?**

RESPONSE: We have now included intra-plate repeatability estimates on lines 251-252. Samples were generally placed next to one another on the plate.

@@

---

\bigskip

**L 247: Please compare cross-sectional and longitudinal telomere shortening and discuss it.**

RESPONSE: We now include a formal test of the within and between individual slopes, as described in van de Pol and Wright (2009), and find that there is no difference between cross-sectional and longitudinal telomere shortening - lines 273-274:

@@

\begin{adjustwidth}{2cm}{2cm}

\textit{"Further, we found no evidence that within and between individual slopes of telomere shortening varied (estimate = -0.003, CIs = -0.008, 0.003), suggesting that there was no difference"}

@@

---

\bigskip

**L 340: Does telomere lengthening or shortening relate to environmental conditions in your study?**

RESPONSE: As stated in the discussion (lines 360-374), we feel that this is an important area for future research. However, this would involve a comprehensive, complex set of new analyses. Difficulties associated with such an analysis are incorporating temporal changes in territory quality/conditions, incorporating the effects of individuals moving territories, and moving beyong detecting telomere lenghthening at a population level, to distinguishing legthening from error at the individual level. Thus, we hope that addressing the relationship between telomere lengthening and enviroinment will form a whole manuscript on its own - but we need to do a lot more work before such a manuscript can be written. From a practical perspective, we are already pushing the word limit for Journal of Animal Ecology, so adding in an entire new set of analyses is just not possible.

@@

---

\bigskip

**Generally, the discussion of this part of potential causes of telomere dynamics remains very methodological. The ecological context could be improved by adding own results as mentioned above or more citations.**

RESPONSE: See response above regarding the analysis of lengthening. Although this section is somewhat methodological, we feel that i) it is a very important point to be able to separate telomere elongation from sampling error, as few studies have done so in ecological systems, and ii) we make it clear (lines 360-374) that our findings are likely to be of ecological importance.
